---
title:  'An analysis of 3D shapes of cities'
subtitle: 'INFO 698: Capstone Project'
author:
- name: Student -  H M Abdul Fattah
  affiliation: Graduate Student, School of Information, University of Arizona, Tucson, AZ
- name: Advisor -  Dr. Cristian Román-Palacios
  affiliation: Assistant Professor, School of Information, University of Arizona, Tucson, AZ
tags: [R, RStudio]
output: html_document
---

# Intall & Load the libraries

```{r message=FALSE}

# Include the libraries that are needed to be installed
libs <- c(
    "tidyverse", "sf", "geodata",
    "terra", "classInt", "rayshader", "dplyr", "utils", 
    "stars", "raster", "giscoR", "tidyterra", "plotly", "gridExtra", "e1071", 
    "diptest", "xtable"
)

installed_libs <- libs %in% rownames(
    installed.packages()
)

if (any(installed_libs == F)) {
    install.packages(
        libs[!installed_libs]
    )
}

invisible(lapply(
    libs,
    library,
    character.only = T
))

```

# Download Cities Shapefile if not present in directory
```{r}
# Define file name and folder path
file_name <- "ne_10m_urban_areas_landscan.shp"
folder_path <- "../data/07_cities_shapefile"

# Check if the folder exists, if not, create it
if (!file.exists(folder_path)) {
  dir.create(folder_path)
}

# Check if the file exists in the folder
if (!file.exists(file.path(folder_path, file_name))) {
  # If not, download the zip file
  url <- "https://stacks.stanford.edu/file/druid:yk247bg4748/data_EPSG_4326.zip"
  download.file(url, destfile = file.path(folder_path, "data_EPSG_4326.zip"))
  
  # Set a timeout of 10 seconds before unzipping
  Sys.sleep(10)
  
  # Unzip the file
  unzip(zipfile = file.path(folder_path, "data_EPSG_4326.zip"), exdir = folder_path)
}
```

# Load the cities shapefile
```{r}
file_name <- "ne_10m_urban_areas_landscan.shp"
folder_path <- "../data/07_cities_shapefile"

# Read the shapefile
world_city_map <- st_read(file.path(folder_path, file_name))
```
# City names stored in a list
```{r}
city_names <- unique(world_city_map$name_conve)
city_names_list <- as.list(city_names)

# Check if a city is present in 'city_names_list'
# 'Tucson' %in% city_names_list
```

# Definition of getContinents() funciton
```{r}
getContinents <- function(city_polys, cont_polys) {
  # Assigns each city to a continent.
  names <- city_polys$name_conve
  continents <- cont_polys$NAME
  cont_list <- rep(NA, nrow(city_polys))
  for (i in 1:nrow(city_polys)) {
    poly <- city_polys$geometry[[i]]
    if (!is.null(poly)) {
      if (!st_is_valid(poly)) {
        poly <- st_buffer(poly, dist = 0)
      }
      center <- st_centroid(poly)
      min_dist <- 999999
      for (j in 1:nrow(cont_polys)) {
        cont_poly <- cont_polys$geometry[[j]]
        d <- st_distance(center, cont_poly)
        if (length(st_intersection(poly, cont_poly)) > 0) {
          cont_list[i] <- continents[j]
          break
        }
        if (d < min_dist) {
          min_dist <- d
          cont_list[i] <- continents[j]
        }
      }
    }
  }
  return(cont_list)
}
```

# Get Continents name for each city
```{r}
# this shape file is already on github. If not: 
# download the "label areas" from:
# url: https://www.naturalearthdata.com/downloads/50m-physical-vectors/50m-physical-labels/
# the downloaded zip file should be inside folder: '09_continents_shapefile'

# Define file name and folder path
file_name <- "ne_50m_geography_regions_polys.shp"
folder_path <- "../data/09_continents_shapefile"

# Check if the file exists in the folder
if (!file.exists(file.path(folder_path, file_name))) {
  # Unzip the file
  unzip(zipfile = file.path(folder_path, "ne_50m_geography_regions_polys.zip"), exdir = folder_path)
}

# Read the shapefile
continents_data <- st_read(file.path(folder_path, file_name))

options(timeout = 5)

# Filter data based on condition
continents_data <- continents_data[continents_data$FEATURECLA == 'Continent', ]

# Call the function
result <- getContinents(world_city_map, continents_data)

options(timeout = 5)

world_city_map$Continent <- result
```

# Download GHSL data

```{r}
# Define file name and folder path
file_name <- "GHS_BUILT_H.zip"
folder_path <- "../data/08_cities_height"

# Check if the folder exists, if not, create it
if (!file.exists(folder_path)) {
  dir.create(folder_path)
}

# Check if the file exists in the folder
if (!file.exists(file.path(folder_path, file_name))) {
  # If not, download the zip file
  url <- "https://jeodpp.jrc.ec.europa.eu/ftp/jrc-opendata/GHSL/GHS_BUILT_H_GLOBE_R2023A/GHS_BUILT_H_AGBH_E2018_GLOBE_R2023A_4326_3ss/V1-0/GHS_BUILT_H_AGBH_E2018_GLOBE_R2023A_4326_3ss_V1_0.zip"
  
  download.file(url, destfile = file.path(folder_path, "GHS_BUILT_H.zip"))
  
  # Set a timeout of 300 seconds before unzipping
  options(timeout = 300)
  
  # Unzip the file
  unzip(zipfile = file.path(folder_path, "GHS_BUILT_H.zip"), exdir = folder_path)
}
```

# Load GHSL data
```{r}
tif_file = "../data/08_cities_height/GHS_BUILT_H_AGBH_E2018_GLOBE_R2023A_4326_3ss_V1_0.tif"

building_h <- terra::rast(tif_file)
```

# Iterate over cities + save png file 
# Warning: Takes time to render for all cities (parameter: n controls the number of cities)
```{r message=FALSE}
counter <- 0
n <- 1 # number of cities

for (city_name in city_names_list) {
  #print(city_name)
  
  city_sf <- world_city_map %>%
    filter(name_conve %in% city_name)
    
  # crop city height from building_h object
  city_height <- terra::crop(
    building_h,
    #convert to terra friendly format
    terra::vect(city_sf), 
    # crop the inside
    snap = "in", 
    # remove everything except city_sf
    mask = T
  )
  
  # Set a timeout of 5 seconds 
  options(timeout = 5)
  
  # terra::aggregate to reduce dim
  city_height_agg <- city_height |>
    terra::aggregate(
        fact = 3
    )
  
  # Output Folder
  output_folder <- "../results/city_rasters/"

  # Check if the output folder exists, if not, create it
  if (!file.exists(output_folder)) {
    dir.create(output_folder, recursive = TRUE)
  }

  # Specify the file path where we want to save the raster
  output_file <- paste(output_folder, city_name, ".tif", sep = "")

  # Write the raster to a GeoTIFF file
  writeRaster(city_height_agg, filename = output_file, overwrite = TRUE)
  
  # Set a timeout of 5 seconds 
  options(timeout = 5)
    
  # convert raster data to dataframe
  city_height_df = as.data.frame(
    city_height, 
    xy = T, 
    na.rm = T
  )
  
  names(city_height_df)[3] <- "height"

  # Print the dimension of the dataframe
  print(dim(city_height_df))
  
  # Print minimum and maximum values of 'height' column in the same line
  cat("Min height:", min(city_height_df$height), ", Max height:", 
      max(city_height_df$height), "\n")
  
  # Calculate breaks and colors
  breaks <- classInt::classIntervals(city_height_df$height, n = 5, 
                                     style = "fisher")$brks
  cols <- c("white", "gray90", "gray70", "gray50", "gray30")
  texture <- colorRampPalette(cols, bias = 2)(6)
  
  # Create a GGPLOT2 object
  plot_each_city <- ggplot(city_height_df) +
    geom_raster(aes(x = x, y = y, fill = height)) +
    scale_fill_gradientn(name = "height (m)", colors = texture, 
                         breaks = round(breaks, 0)) +
    coord_sf(crs = 4326) +
    guides(fill = guide_legend(direction = "vertical", keyheight = unit(5, "mm"),
                               keywidth = unit(5, "mm"), title.position = "top", 
                               label.position = "right", title.hjust = .5, 
                               label.hjust = .5, ncol = 1, byrow = FALSE)) +
    theme_minimal() +
    theme(axis.line = element_blank(), axis.title.x = element_blank(), 
          axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), 
          legend.position = "right",
          legend.title = element_text(size = 11, color = "grey10"),
          legend.text = element_text(size = 10, color = "grey10"),
          panel.grid.major = element_line(color = "white"),
          panel.grid.minor = element_line(color = "white"),
          plot.background = element_rect(fill = "white", color = NA),
          legend.background = element_rect(fill = "white", color = NA),
          panel.border = element_rect(fill = NA, color = "white"),
          plot.margin = unit(c(t = 0, r = 0, b = 0, l = 0), "lines"))
  
  # Plot the scene using rayshader
  rayshader::plot_gg(
    ggobj = plot_each_city,
    width = 4,
    height = 7,
    scale = 150,
    solid = FALSE,
    soliddepth = 0,
    shadow = TRUE,
    shadow_intensity = 0.99,
    shadowdepth = 0,
    offset_edges = FALSE,
    sunangle = 315,
    window.size = c(1000, 800),
    zoom = 0.44,
    phi = 30,
    theta = 0, # rotation angle
    multicore = TRUE
  )
  
  # Output folder
  out_folder = "../results/figures/"
  
  # Low quality image
  # Render the snapshot directly to a PNG file
  
  #rayshader::render_snapshot(
  #  filename = paste0(out_folder, city_name, "-3D-map.png"),
  #  width = 3000,
  #  height = 3000
  #)
  
  # High Quality image
  # Render the scene 
  rayshader::render_highquality(
    filename = paste0(out_folder, city_name, "-3D-map-2018.png"),
    preview = TRUE,
    interactive = FALSE,
    light = TRUE,
    lightdirection = c(315, 310, 315, 310),
    lightintensity = c(1000, 1500, 150, 100),
    lightaltitude = c(15, 15, 80, 80),
    ground_material = rayrender::microfacet(roughness = 0.6),
    width = 700,
    height = 700
  )
  
  # Close the plot
  rgl::close3d()
  
  counter <- counter + 1
  if (counter >= n) {
    break  # Break the loop when counter reaches n
  }
}
```

# Density plot
# X axis is building heights
# Y axis is the frequency of area (area is calculated by cellSize function)
# [Simiar to height vs area]
# Note: I am keeping this chunk of code for reference or future use
```{r}

counter <- 0
n <- 5 # number of cities

for (city_name in city_names_list) {
  #print(city_name)
  
  city_sf <- world_city_map %>%
    filter(name_conve %in% city_name)
    
  # crop city height from building_h object
  city_height <- terra::crop(
    building_h,
    #convert to terra friendly format
    terra::vect(city_sf), 
    # crop the inside
    snap = "in", 
    # remove everything except city_sf
    mask = T
  )
  
  # convert raster data to dataframe
  city_height_df = as.data.frame(
    city_height, 
    xy = T, 
    na.rm = T
  )
  
  names(city_height_df)[3] <- "height"
  
  # Calculate area covered by each cell in city_height raster
  cell_size <- cellSize(city_height)
  
  city_area_df = as.data.frame(
    cell_size, 
    xy = T, 
    na.rm = T
  )
  
  # Merge city_height_df and city_area_df based on x and y coordinates
  merged_df <- merge(city_height_df, city_area_df, by = c("x", "y"))
  
  # Remove 'x' and 'y' columns from merged dataframe
  new_df <- merged_df[, !(names(merged_df) %in% c("x", "y"))]
  
  # Create a KDE plot
  p <- ggplot(new_df, aes(x = height)) +
    geom_density(fill = "skyblue", color = "black") +
    ggtitle("Density Plot of Building Heights") +
    xlab("Height") +
    ylab("Density") +
    theme_minimal() +
    theme(panel.grid = element_blank())
  
  #print(p)
  
  # Compute density estimates
  density <- density(new_df$height)
  
  # Reverse the sign of density values
  density$y <- -density$y
  
  # Create a KDE plot with density values as negative
  p <- ggplot(data.frame(x = density$x, y = density$y), aes(x, y)) +
    geom_area(fill = "skyblue", color = "black", alpha = 0.5) +
    ggtitle("Density Plot of Building Heights (Mirrored)") +
    xlab("Height") +
    ylab("Density (Negative)") +
    theme_minimal()
  
  #print(p)
  
  new_df$height <- round(new_df$height, 2)
  
  # Compute density estimates for positive and negative values
  density_positive <- density(new_df$height)
  density_negative <- density(new_df$height)
  
  # Reverse the sign of the negative density values
  density_negative$y <- -density_negative$y
  
  # Combine density values into a single dataset
  density_combined <- data.frame(
    height = density_positive$x,
    positive_density = density_positive$y,
    negative_density = density_negative$y
  )
  
  # Create a plot of positive and negative density values with respect to height
  p <- ggplot(density_combined, aes(x = height)) +
    geom_area(aes(y = positive_density, fill = "Positive"), color = "skyblue", linetype = "solid") +
    geom_area(aes(y = negative_density, fill = "Negative"), color = "skyblue", linetype = "solid") +
    ggtitle(paste("Area vs Building Heights for city :", city_name)) +
    theme_minimal() +
    coord_flip() +
    scale_fill_manual(values = c("Positive" = "skyblue", "Negative" = "skyblue")) + 
    guides(fill = FALSE) +  # Hide the legend
    theme(
      axis.title.x = element_blank(),  # Remove x-axis title
      axis.title.y = element_blank(),  # Remove y-axis title
      panel.grid.major.x = element_blank(),  # Remove major grid lines on x-axis
      panel.grid.minor.x = element_blank(),  # Remove minor grid lines on x-axis
      panel.grid.major.y = element_blank(),  # Remove major grid lines on y-axis
      panel.grid.minor.y = element_blank()  # Remove minor grid lines on y-axis
    )
  
  print(p)
  
  # Remove leading and trailing whitespace from the city name
  city_name <- trimws(city_name)

  # Save the plot as a .png file without labels or background
  #ggsave(paste("../results/figures/", city_name, "-area-map.png"), plot = p, width = 8, height = 6, units = "in", dpi = 300, bg = "transparent")
  
  counter <- counter + 1
  if (counter >= n) {
    break  # Break the loop when counter reaches n
  }
}

```

# Draw the city using plotly
# Just a simple 3D represenation of building heights. This plot is interactive 
```{r}

counter <- 0
n <- 1

for (city_name in city_names_list) {
  city_sf <- world_city_map %>%
    filter(name_conve %in% city_name)
  
  city_height <- terra::crop(
    building_h,
    terra::vect(city_sf), 
    snap = "in", 
    mask = TRUE
  )
  
  city_height_df <- as.data.frame(city_height, xy = TRUE, na.rm = TRUE)
  names(city_height_df)[3] <- "height"
  
  # Convert 'height' column to numeric
  city_height_df$height <- as.numeric(as.character(city_height_df$height))
  
  # Get unique x and y values
  unique_x <- sort(unique(city_height_df$x))
  unique_y <- sort(unique(city_height_df$y))
  
  # Create a matrix to store height values
  city_height_matrix <- matrix(NA, nrow = length(unique_x), ncol = length(unique_y))
  
  # Fill the matrix with height values
  for (i in 1:length(unique_x)) {
    for (j in 1:length(unique_y)) {
      city_height_matrix[i, j] <- mean(city_height_df$height[city_height_df$x == unique_x[i] & city_height_df$y == unique_y[j]])
    }
  }
  
  # Create a 3D surface plot
  p2 <- plot_ly(z = ~city_height_matrix, type = "surface") %>%
    layout(scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Height")
    ))
  
  print(p2)
  
  counter <- counter + 1
  if (counter >= n) {
    break  # Break the loop when counter reaches n
  }
}
```


```{r}

```

# Calculate mean, skewness, modality, classificaiton label
```{r}
counter <- 0
n <- 2 # number of cities

Pyramid_Count <- 0
Inverse_Pyramid_Count <- 0
Diamond_Count <- 0
Hourglass_Count <- 0
Unknown_Count <- 0

# Create an empty list to store tables
table_list <- list()

for (city_name in city_names_list) {
  city_sf <- world_city_map %>%
    filter(name_conve %in% city_name)
  
  # crop city height from building_h object
  city_height <- terra::crop(
    building_h,
    terra::vect(city_sf), 
    snap = "in", 
    mask = T
  )
  
  # convert raster data to dataframe
  city_height_df = as.data.frame(
    city_height, 
    xy = T, 
    na.rm = T
  )
  
  names(city_height_df)[3] <- "height"
  
  # Calculate area covered by each cell in city_height raster
  cell_size <- cellSize(city_height)
  
  city_area_df = as.data.frame(
    cell_size, 
    xy = T, 
    na.rm = T
  )
  
  # Merge city_height_df and city_area_df based on x and y coordinates
  merged_df <- merge(city_height_df, city_area_df, by = c("x", "y"))
  
  # Remove 'x' and 'y' columns from merged dataframe
  area_vs_height_data <- merged_df[, !(names(merged_df) %in% c("x", "y"))]
  
  # Calculate mean height
  mean_height <- mean(area_vs_height_data$height, na.rm = TRUE)
  
  # Calculate skewness
  skew <- skewness(area_vs_height_data$height)
  #print(paste("Skewness: ", skew))
  
  # Calculate dip statistic
  dip <- dip.test(area_vs_height_data$height)$p.value
  #print(paste("Dip Statistic: ", dip))
  
  # Calculate modality
  modality <- ifelse(dip > 0.001, "Unimodal ", "Multimodal ")
  #print(paste("Modality: ", modality))
  
  # Continent 
  city_df <- world_city_map[world_city_map$name_conve == city_name, ]
  city_continent_name <- city_df$Continent
  
  # Calculate Class
  class <- ""
  
  # Check if skew is not NA
  if (!is.na(skew)) {
    # Calculate Class
    class <- ""
    if (dip > 0.001)
    {
      if (skew > 0) {
        # Unimodal right skew
        class <- "Pyramid"
        Pyramid_Count <- Pyramid_Count + 1
      } else if (skew < 0) {
        # Unimodal left skew
        class <- "Inverse Pyramid" 
        Inverse_Pyramid_Count <- Inverse_Pyramid_Count + 1
      } else {
        # Unimodal with no skew
        class <- "Diamond"
        Diamond_Count <- Diamond_Count + 1
      }
    }
    else
    {
      class <- "Hourglass"
      Hourglass_Count <- Hourglass_Count + 1
    }
  } else {
    # If skew is NA, assign appropriate values
    skew <- 0
    modality <- "Unknown"
    class <- "Unknown"
    Unknown_Count <- Unknown_Count + 1
  }
  
  # Round the height values by 2 decimals
  area_vs_height_data$height <- round(area_vs_height_data$height, 1)
  
  # Convert area from m^2 to km^2, conversion: 1 square kilometer = 1,000,000 square meters
  area_vs_height_data$area <- area_vs_height_data$area / 1000000  
  
  # Aggregate the areas at each unique rounded height value
  aggregated_area_data <- aggregate(area ~ height, data = area_vs_height_data, FUN = sum)

  p <- ggplot(aggregated_area_data) + 
    geom_area(aes(y = area, x = height)) +
    theme_minimal() +
    xlab("Height (m)") +  # Modify x-label
    ylab("Area (km"^"2"~")") +    # 
    geom_vline(xintercept = mean_height, linetype = "dotted", color = "green", size = 1.5)
  
  # Create legend text
  legend_text <- paste(city_continent_name,
                       "\ncity name = ", city_name,
                       "\nclass = ", class,
                       "\nskew = ", round(skew, 2),
                       "\ndip = ", round(dip, 2),
                       "\nmean = ", round(mean_height, 2),
                       "\n", modality)
  
  # Add annotation for legend
  p <- p + annotate("text", x = Inf, y = Inf, label = legend_text, 
                    hjust = 1, vjust = 1, size = 5, color = "black")
  
  p <- p + theme(panel.grid = element_blank())
  
  print(p)
  
  #ggsave(paste("../results/figures/", city_name, "-plot-a.png"), plot = p, width = 8, height = 6, units = "in", dpi = 300, bg="white")
  
  # Combine density values into a single dataset
  area_combined <- data.frame(
    height = aggregated_area_data$height,
    positive_area = aggregated_area_data$area,
    negative_area = -1 * aggregated_area_data$area
  )
  
  # Create a plot of positive and negative density values with respect to height
  p <- ggplot(area_combined, aes(x = height)) +
    geom_area(aes(y = positive_area, fill = "Positive"), color = "skyblue", linetype = "solid") +
    geom_area(aes(y = negative_area, fill = "Negative"), color = "skyblue", linetype = "solid") +
    ggtitle(paste("Density Plot of Building Heights for city : ", city_name)) +
    theme_minimal() +
    coord_flip() +
    scale_fill_manual(values = c("Positive" = "skyblue", "Negative" = "skyblue")) + 
    guides(fill = FALSE) +  # Hide the legend
    theme(
      axis.title.x = element_blank(),  # Remove x-axis title
      axis.title.y = element_blank(),  # Remove y-axis title
      panel.grid.major.x = element_blank(),  # Remove major grid lines on x-axis
      panel.grid.minor.x = element_blank(),  # Remove minor grid lines on x-axis
      panel.grid.major.y = element_blank(),  # Remove major grid lines on y-axis
      panel.grid.minor.y = element_blank()  # Remove minor grid lines on y-axis
    )
  
  print(p)
  
  # Remove leading and trailing whitespace from the city name
  city_name <- trimws(city_name)
  
  # Create a dataframe with the summary information
  summary_df <- data.frame(
    City = city_name,
    Continent = city_continent_name,
    Class = class,
    Skewness = skew,
    Dip_Statistic = dip,
    Mean_Height = mean_height
  )
  
  # Save the plot as a .png file without labels or background
  #bg = "transparent"
  ggsave(paste("../results/figures/", city_name, "-area-vs-height-map.png"), plot = p, width = 8, height = 6, units = "in", dpi = 300, bg = "white")
  
  # Add the summary dataframe to the list for generating pdf later on
  table_list[[counter + 1]] <- summary_df
  
  counter <- counter + 1
  if (counter >= n) {
    break  # Break the loop when counter reaches n
  }
}


# Combine all tables into one dataframe
combined_table <- do.call(rbind, table_list)

print('Classification Type Count: ')
print(Pyramid_Count)
print(Inverse_Pyramid_Count)
print(Diamond_Count)
print(Hourglass_Count)
print(Unknown_Count)

```

# Save to CSV
```{r}
# Define the file path for the CSV
csv_file <- "../docs/combined_summary.csv"

# Write the combined table to a CSV file
write.csv(combined_table, file = csv_file, row.names = FALSE)

# Confirm that the CSV file has been saved
if (file.exists(csv_file)) {
  message("CSV file successfully generated: ", csv_file)
} else {
  warning("CSV file generation failed.")
}
```


```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```